---
title: "Bluefish index standardization - NJ Ocean trawl"
author: "M Celestino"
date: "`r Sys.time()`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = FALSE)
```

## Bluefish index standardization

### Step 1: Data processing

Load functions & libraries
```{r}
#source("f:/BF_WG/FIG/figFuns.R")
root <- "C:/Users"
usr <- "jgorzo"
loc <- "OneDrive - New Jersey Office of Information Technology/Documents"
root <- file.path(root, usr, loc)
source(file.path(root, "data/tog/bootstrap_functions.R"))

library(car, quietly = FALSE)
library(lmtest, quietly = FALSE)
library(glmmTMB, quietly = FALSE)
library(DHARMa, quietly = FALSE)
library(bbmle, quietly = FALSE)
library(ggeffects, quietly = FALSE)
library(ggplot2, quietly = FALSE)
require(readxl, quietly = FALSE)
require(lattice, quietly = FALSE)
require(emmeans, quietly = FALSE)
require(rmarkdown, quietly = FALSE)
require(knitr, quietly = FALSE)
library(reshape2, quietly=TRUE, verbose=FALSE)
library(doBy, quietly=TRUE, verbose=FALSE)
library(dplyr, quietly=TRUE, verbose=FALSE,warn.conflicts=FALSE)

geomean0 <- function(x) {
  exp(mean(log(x + 1))) - 1
}

pctPos <- function(x) {
  x <- ifelse(x > 0, 1, 0)
  mean(x)
}

disp3 <- function(mod = NA) {
  E <- residuals(mod, type = "pearson")
  d <- sum(E^2) / summary(mod)$AICtab[5]
  return(d)
}
```

### Step 2: Model selection 

Compare negative binom, zero-infl neg binom, zero-altered neg binom

#### Model 1: Negative Binomial
```{r,error=TRUE}
tmb1.NB = glmmTMB(mod, offset=lnEffort, data = dat, family = nbinom2)
summary(tmb1.NB)
# Calculate dispersion for a glmmTMB model
disp3(tmb1.NB)
```

#### Model 2: ZINB
```{r, error=TRUE, eval=F}
#ZINB = glmmTMB(mod,offset=lnEffort, 
#               ziformula = ~ YEAR + STRATA + DEPTH + BDO + BTEMP, data = dat, 
#               control=glmmTMBControl(optCtrl=list(iter.max=5e6,eval.max=5e6)),
#               family=nbinom2) # convergence problems
#summary(ZINB)

pairs(~ YEAR + STRATA + DEPTH + BDO + BTEMP,data=dat) # also looks like non-linear relationship among DEPTH and BSAL and CPUE
```

```{r}
# Previous model bombs; large yr estimates in ZI model, so remove YEAR
ZINB <- glmmTMB(mod,offset=lnEffort, 
                ziformula = ~ STRATA + DEPTH + BDO  + BTEMP, data = dat, 
                control=glmmTMBControl(optCtrl=list(iter.max=5e3,eval.max=5e3)),
                family=nbinom2)
summary(ZINB) # THIS WORKS FOR TOG- JESS
residuals(ZINB, type = "pearson")
#tmbZI.resids(ZINB, y.dat=dat$CPUE)

#JESS DOESNT NEED THIS
# large SEs in strata in ZI model, so removing
#ZINB <- glmmTMB(CPUE ~ YEAR + STRATA + DEPTH + BDO,offset=lnEffort, 
#                ziformula = ~ DEPTH + BDO, data = dat, 
#                control=glmmTMBControl(optCtrl=list(iter.max=5e3,eval.max=5e3)),
#                family=nbinom2)
#summary(ZINB) # ok
```

#### Model 3: ZANB (aka hurdle model)
```{r, eval=F, error=TRUE}
ZANB <- glmmTMB(mod, 
                ziformula = ~ YEAR + STRATA + DEPTH + BDO + BTEMP, data = dat,
                offset=lnEffort,family = truncated_nbinom2(link = "log"))
summary(ZANB) #convergence problem
```

```{r eval=F}
# high SEs in YEAR, so removing
ZANB <- glmmTMB(mod, 
                ziformula = ~ DEPTH + BDO + BTEMP, data = dat,
                offset=lnEffort,family = truncated_nbinom2(link = "log"))
summary(ZANB) # doesn't converge

#tmbZI.resids(ZANB, y.dat=dat$CPUE)
```


#### Preliminary model comparisons
```{r, error=TRUE}
AICtab(tmb1.NB, ZINB) # ZANB won for bluefish but won't converge for tog
```


### Step 3: Factor selection (by model)

#### Negative Binomial
```{r, error=TRUE}
NB1 <- glmmTMB(mod, offset=lnEffort, data = dat, family = nbinom2)
#summary(NB1)
drop1(NB1,test="Chi")

# drop BDO:
NB2 <- glmmTMB(CPUE ~ YEAR + STRATA + DEPTH + offset(lnEffort), data = dat, family = nbinom2)
#summary(NB2)
drop1(NB2,test="Chi")

# drop depth:
NB3 <- glmmTMB(CPUE ~ YEAR + STRATA + offset(lnEffort), data = dat, family = nbinom2)
#summary(NB3)
drop1(NB3,test="Chi") # all terms significant

# Let's compare to a model with YEAR only.
NB0 <- glmmTMB(CPUE ~ YEAR+offset(lnEffort), data = dat, family = nbinom2)
#summary(NB0)
```

```{r}
NB5 <- glmmTMB(CPUE ~ YEAR + STRATA + DEPTH + BDO, offset=lnEffort, data = dat, family = nbinom2)
#summary(NB5)
drop1(NB5,test="Chi")

NB6 <- glmmTMB(CPUE ~ YEAR + STRATA + DEPTH + BTEMP, offset=lnEffort, data = dat, family = nbinom2)
#summary(NB6)
drop1(NB6,test="Chi")

NB7 <- glmmTMB(CPUE ~ YEAR + STRATA + BTEMP, offset=lnEffort, data = dat, family = nbinom2)
#summary(NB7)
drop1(NB7,test="Chi")

NB8 <- glmmTMB(CPUE ~ YEAR + BTEMP, offset=lnEffort, data = dat, family = nbinom2)
#summary(NB8)
drop1(NB8,test="Chi")

NB9 <- glmmTMB(CPUE ~ YEAR + BDO, offset=lnEffort, data = dat, family = nbinom2)
#summary(NB8)
drop1(NB9,test="Chi")

NB10 <- glmmTMB(CPUE ~ YEAR + STRATA + BDO, offset=lnEffort, data = dat, family = nbinom2)
#summary(NB8)
drop1(NB10,test="Chi")

NB11 <- glmmTMB(CPUE ~ YEAR + STRATA + BDO + BTEMP, offset=lnEffort, data = dat, family = nbinom2)
#summary(NB8)
drop1(NB11,test="Chi")

NB12 <- glmmTMB(CPUE ~ YEAR + BDO + BTEMP, offset=lnEffort, data = dat, family = nbinom2)
#summary(NB8)
drop1(NB12,test="Chi")

NB13 <- glmmTMB(CPUE ~ YEAR + DEPTH + BDO + BTEMP, offset=lnEffort, data = dat, family = nbinom2)
#summary(NB8)
drop1(NB13,test="Chi")
```

```{r}
NB14 <- glmmTMB(CPUE ~ YEAR + DEPTH + BTEMP, offset=lnEffort, data = dat, family = nbinom2)
#summary(NB8)
drop1(NB14,test="Chi")

AICtab(NB1, NB2, NB3, NB0, NB5, NB6, NB7, NB8, NB9, NB10, NB12, NB11, NB13, NB14) # so NB1 wins
lrtest(NB0, NB3, NB2, NB1)
sapply(list(NB3,NB2,NB1,NB0),disp3)
```


#### ZINB
```{r, error=TRUE}
# ZINB (starting with final model from above):
#ZINB1 <- glmmTMB(mod,offset=lnEffort, 
#                ziformula = ~ STRATA + YEAR + DEPTH + BDO + BTEMP, data = dat, 
#                control=glmmTMBControl(optCtrl=list(iter.max=5e3,eval.max=5e3)),
#                family=nbinom2)
#summary(ZINB1)
#drop1(ZINB1,test="Chi") 

# ZINB (starting with final model from above):
ZINB1 <- glmmTMB(mod,offset=lnEffort, 
                ziformula = ~ DEPTH + BDO + BTEMP, data = dat, 
                control=glmmTMBControl(optCtrl=list(iter.max=5e3,eval.max=5e3)),
                family=nbinom2)
summary(ZINB1)
drop1(ZINB1,test="Chi") # convergence problems
```

```{r}
# drop DO from conditional (not significant):
ZINB2 <- glmmTMB(CPUE ~ YEAR + STRATA + DEPTH,offset=lnEffort, 
                ziformula = ~ STRATA + DEPTH, data = dat, 
                control=glmmTMBControl(optCtrl=list(iter.max=5e3,eval.max=5e3)),
                family=nbinom2) # convergence problems
summary(ZINB2)
drop1(ZINB2,test="Chi")
```

```{r}
#CPUE ~ YEAR + STRATA + DEPTH + BTEMP
# drop depth from ZI (very param large estimate):
ZINB3 <- glmmTMB(CPUE ~ YEAR + STRATA + DEPTH + BTEMP,offset=lnEffort, 
                ziformula = ~ STRATA + DEPTH + BTEMP, data = dat, 
                control=glmmTMBControl(optCtrl=list(iter.max=5e3,eval.max=5e3)),
                family=nbinom2) # convergence problems
#summary(ZINB3)
#drop1(ZINB3,test="Chi")

ZINB6 <- glmmTMB(CPUE ~ YEAR + STRATA + DEPTH + BTEMP,
                offset=lnEffort, 
                ziformula = ~ STRATA + DEPTH, data = dat, 
                control=glmmTMBControl(optCtrl=list(iter.max=5e3,eval.max=5e3)),
                family=nbinom2) # convergence problems


ZINB7 <- glmmTMB(CPUE ~ YEAR + STRATA + DEPTH + BTEMP,
                offset=lnEffort, 
                ziformula = ~ STRATA, data = dat, 
                control=glmmTMBControl(optCtrl=list(iter.max=5e3,eval.max=5e3)),
                family=nbinom2) # convergence problems
```

```{r}
ZINB8 <- glmmTMB(CPUE ~ YEAR + STRATA + DEPTH + BTEMP,
                offset=lnEffort, 
                ziformula = ~ DEPTH, data = dat, 
                control=glmmTMBControl(optCtrl=list(iter.max=5e3,eval.max=5e3)),
                family=nbinom2) # convergence problems
#summary(ZINB3)
#drop1(ZINB3,test="Chi")
```

```{r}
# drop BDO from ZI (now no ZI model):
ZINB4 <- glmmTMB(CPUE ~ YEAR + STRATA + DEPTH,
                offset=lnEffort, 
                data = dat, 
                control=glmmTMBControl(optCtrl=list(iter.max=5e3,eval.max=5e3)),
                family=nbinom2)
summary(ZINB4)
drop1(ZINB4,test="Chi")
```

```{r}
# drop depth:
ZINB5 <- glmmTMB(CPUE ~ YEAR + STRATA,offset=lnEffort, 
                data = dat, 
                control=glmmTMBControl(optCtrl=list(iter.max=5e3,eval.max=5e3)),
                family=nbinom2)
summary(ZINB5)
drop1(ZINB5,test="Chi") # all significant


# a model with YEAR only:
ZINB0 <- glmmTMB(CPUE ~ YEAR,offset=lnEffort, 
                data = dat, 
                control=glmmTMBControl(optCtrl=list(iter.max=5e3,eval.max=5e3)),
                family=nbinom2)
summary(ZINB0)
```

```{r}
ZINB9 <- glmmTMB(CPUE ~ YEAR + STRATA + DEPTH + BTEMP,
                offset=lnEffort, 
                ziformula = ~ BTEMP, data = dat, 
                control=glmmTMBControl(optCtrl=list(iter.max=5e3,eval.max=5e3)),
                family=nbinom2) # convergence problems
#summary(ZINB3)
#drop1
```

```{r}
ZINB10 <- glmmTMB(CPUE ~ YEAR + STRATA + DEPTH + BTEMP,
                offset=lnEffort, 
                ziformula = ~ STRATA + BTEMP, data = dat, 
                control=glmmTMBControl(optCtrl=list(iter.max=5e3,eval.max=5e3)),
                family=nbinom2) # convergence problems
#summary(ZINB3)
#drop1
```

```{r}
ZINB11 <- glmmTMB(CPUE ~ YEAR + STRATA + DEPTH + BTEMP,
                offset=lnEffort, 
                ziformula = ~ DEPTH + BTEMP, data = dat, 
                control=glmmTMBControl(optCtrl=list(iter.max=5e3,eval.max=5e3)),
                family=nbinom2) # convergence problems
#summary(ZINB3)
#drop1
```

```{r}
AICtab(ZINB5,ZINB0, ZINB1, ZINB2, ZINB3, ZINB4, ZINB6, ZINB7, ZINB8, ZINB9, ZINB10, ZINB11) # bluefish: ZINB1-3 had convergence problems so not including
lrtest(ZINB0,ZINB5, ZINB1,ZINB2, ZINB3, ZINB4)
#sapply(list(ZINB5,ZINB4,ZINB0),tmbZI.resids,dat$CPUE)
```

#### ZANB (aka hurdle model)
```{r, eval=F, error=TRUE}
# ZANB - starting with final model from model selection above
ZANB1 <- glmmTMB(mod, 
   ziformula = ~ STRATA + BTEMP, 
   data = dat,
   offset=lnEffort,
   family = truncated_nbinom2(link = "log"))
summary(ZANB1)
drop1(ZANB1,test="Chi")
```

```{r eval=FALSE}
# drop BDO from conditional:
ZANB2 <- glmmTMB(CPUE ~ YEAR + DEPTH + BDO + BTEMP, 
   ziformula = ~ BTEMP + BDO, data = dat,
   offset=lnEffort,family = truncated_nbinom2(link = "log"))
summary(ZANB2)
drop1(ZANB2,test="Chi")
```

```{r eval=FALSE}
# drop Depth from conditional:
ZANB3 <- glmmTMB(CPUE ~ YEAR + STRATA + BDO + BTEMP, 
   ziformula = ~ STRATA, data = dat,
   offset=lnEffort,family = truncated_nbinom2(link = "log"))
summary(ZANB3)
drop1(ZANB3,test="Chi")
```

```{r eval=FALSE}
# remove BDO from ZI:
ZANB4 <- glmmTMB(CPUE ~ YEAR + STRATA + DEPTH + BTEMP, 
   ziformula = ~ DEPTH + STRATA, data = dat,
   offset=lnEffort,family = truncated_nbinom2(link = "log"))
summary(ZANB4)
drop1(ZANB4,test="Chi")
```

```{r eval=FALSE}
# remove depth from ZI:
ZANB5 <- glmmTMB(CPUE ~ YEAR + STRATA + DEPTH + BDO, 
   ziformula =  ~ YEAR + STRATA + DEPTH + BDO, data = dat,
   offset=lnEffort,family = truncated_nbinom2(link = "log"))
summary(ZANB5)
drop1(ZANB5,test="Chi") # all significant
```

```{r eval = F}
# a YEAR-only model:
# remove BDO from ZI:
ZANB0 <- glmmTMB(CPUE ~ YEAR + STRATA + DEPTH, 
   ziformula = ~ DEPTH, data = dat,
   offset=lnEffort,family = truncated_nbinom2(link = "log"))
summary(ZANB0) # nope
```

```{r eval=FALSE}
# try again:
ZANB0 <- glmmTMB(CPUE ~ YEAR + STRATA + BDO, 
   ziformula = ~ YEAR + BDO,
   data = dat,
   offset=lnEffort,family = truncated_nbinom2(link = "log"))
   # nope; errors

# model comparison:
AICtab(ZANB1, ZANB2, ZANB3, ZANB4, ZANB5)
lrtest(ZANB5, ZANB4, ZANB3, ZANB2, ZANB1)
#sapply(list(ZANB5,ZANB4,ZANB3,ZANB2,ZANB1),tmbZI.resids,dat$CPUE)

### Model comparisons

AICtab(NB3,ZINB5)

### Model diagnostics

#### All models

# Attempts at some remedial measures:
# all inshore strata seem to have sufficiently high pct pos tows, so not much to do here
op <- par(mfrow=c(1,1))
barplot(with(dat, tapply(CPUE,STRATA,pctPos)),las=3,main="Pct pos by strata") 


# try re-assinging large catch to the 95th pctl of catch (didn't help):
.all <- dat
# dat <- .all  # nrow=341
dat$CPUE[which.max(dat$CPUE)] <- quantile(dat$CPUE,0.95)

ZANB5.1 <- glmmTMB(CPUE ~ YEAR + STRATA, 
   ziformula = ~ STRATA, data = dat,
   offset=lnEffort,family = truncated_nbinom2(link = "log"))
summary(ZANB5.1)
drop1(ZANB5.1,test="Chisq")
op <- par(mfrow=c(1,2))
plotResiduals(simulateResiduals(ZANB5.1),rank=FALSE) 
plotResiduals(simulateResiduals(ZANB5.1),rank=TRUE)
# low residuals at high model predictions, esp predictions over ~500 (which is near the 95th pctl of catches)
# so, I'll re-assigning those values to the 95th percentile of timeseries

dat <- .all
dat$CPUE[dat$CPUE > quantile(dat$CPUE,0.95)] <- quantile(dat$CPUE,0.95)
ZANB5.2 <- glmmTMB(CPUE ~ YEAR + STRATA, 
   ziformula = ~ STRATA, data = dat,
   offset=lnEffort,family = truncated_nbinom2(link = "log"))
summary(ZANB5.2)
drop1(ZANB5.2,test="Chisq")
op <- par(mfrow=c(1,2))
plotResiduals(simulateResiduals(ZANB5.2),rank=FALSE) 
plotResiduals(simulateResiduals(ZANB5.2),rank=TRUE)
# didn't really help, so going back to ZANB5

dat <- .all


# model diagnostics:
models <- list(ZANB5,NB3,ZINB5)
for(i in 1:length(models)) {
   NB.sim <- simulateResiduals(models[[i]])
   this <- names(dat)[names(dat) %in% names(models[[i]]$frame) & names(dat) != "CPUE"]
   print(this)
   plot(NB.sim); mtext(paste(this,collapse=" + "),side=3,adj=0,outer=FALSE,line=1,col="blue")

   for(j in this) {
      plotResiduals(NB.sim, form=dat[,j]); mtext(j,3,col="blue",adj=1)
   }
}


op <- par(mfrow=c(1,2))
plotResiduals(simulateResiduals(ZANB5),rank=FALSE) 
plotResiduals(simulateResiduals(ZANB5),rank=TRUE)
### Model comparisons redux
AICtab(ZANB5,ZANB5.1,ZANB5.2,NB3,ZINB5)

#Conclusion: ZANB5.2 (re-assign all catches > 95th pctl to the 95th pctl) has the lowest AIC, but diagnostics weren't much better than those from ZANB5, and the re-assignment seems like a pretty hefty change. ZANB5 diagnostics were perhaps the same to a tiny bit better than NB3 and ZINB5 (and recall that ZINB5 had no ZI model due to convergence and large SE issues). And, as the last plot above shows, the model has low DHARMa residuals at high predictions; otherwise, residuals are ok. This could be good support for nominal index rather than model-based index.

### Generate and compare indices
# quick comparison of index generated via ZANB5 vs ZANB5.2 (not very different)
outa <- byHand2(ZANB5,nBoot=2,Method="ZANB5", modelType = "zero")
outb <- byHand2(ZANB5.2,nBoot=2,Method="ZANB5.2", modelType = "zero")
matplot(outa$YEAR,data.frame(scaleToMean(outa$Index),scaleToMean(outb$Index)),
   type="o",xlab="",ylab="Index scaled to mean",main="Indicies scaled to mean")
cor(outa$Index,outb$Index)
cor(outa$Index,outb$Index,method="spearman")
```

```{r}
# generate actual final index:
#out <- boot.ZI(ZINB) low convergence 26.2% of bootstrap runs converged.
#out3 <- boot.ZI(ZINB3) #low convergence 26.2% of bootstrap runs converged.
#out4 <- boot.ZI(ZINB2) #"34.9% of bootstrap runs converged."
#out2 <- boot.NB(NB1) # 45.4% of bootstrap runs converged.
#out6 <- boot.ZI(ZINB6)
#out <- byHand2(ZINB,nBoot=1000,Method="ZINB", modelType = "zero")


#out <- boot.NB(NB6) #"43.8% of bootstrap runs converged." best model considered....
#out7 <- boot.ZI(ZANB00)

ZINB00 <- glmmTMB(CPUE ~ 1 + YEAR + STRATA + BDO + BTEMP, offset=lnEffort, data = dat, 
                  ziformula = ~STRATA + BTEMP + BDO,
                  family = nbinom2)
#Zout00 <- boot.NB(ZINB00) #best yet...44.7%

```

Summary:
ZANB5.2 (re-assign all catches > 95th pctl to the 95th pctl) had the lowest AIC, but diagnostics weren't much better than those from ZANB5, and the re-assignment seems like a pretty hefty change. ZANB5 diagnostics were perhaps the same to a tiny bit better than NB3 and ZINB5 (and recall that ZINB5 had no ZI model due to convergence and large SE issues). ZINB5 
  (and alikes) has low DHARMa residuals at high predictions; otherwise, residuals are ok. This could be good support for nominal index rather than model-based index.